---
title: "IV_modeling_2SLS_GMM"
author: "Ileana Marroquin"
date: "2023-10-13"
output:
 html_document:
   toc: true
   toc_depth: 3
   theme: united
   highlight: tango
   css: my.css
---

# Thesis overview

------------------------------------------------------------------------

**Topic: Remittances Vs Public Finance Capacity. Impact Assessment in Reducing Income Inequality in Latin America: A Quasi-Experimental Analysis**

**Abstract:** *Income inequality is a major challenge in Latin America, and remittances are often seen as a potential solution. However, the impact of remittances on income inequality is complex and contested. This thesis examines the impact of remittances on income inequality in Latin America at a macroeconomic level. It utilizes a 15- country, 20-year panel dataset sourced from the World Inequality Database and the World Income Inequality Database. To address endogeneity concerns, this study employs lagged remittances at 1 and 3-year intervals as instrumental variables. Both empirical approaches, namely 2SLS and GMM, indicate that remittances exhibit a weak and statistically insignificant effect on reducing income inequality in the absence of systematic fiscal redistribution. In contrast, public finance capacity demonstrates a statistically significant and more pronounced impact on reducing income inequality. These findings suggest that public finance capacity plays a pivotal role in reducing the Gini index in Latin America surpassing the impact of remittances. Therefore, policymakers should prioritize enhancing public finance capacity to alleviate income inequality.*

**Study Hypothesis:**

-   Ho1: Remittances do not reduce economic inequality

-   Ho2: Remittances sent by migrants to LAC countries are less likely than government expenditure to reduce income inequality

**Input:** the "database_for_IV_modeling.csv" processed database

**Output:** Tables from the modeling and validity test relevant for Chapter 5 of the Thesis

------------------------------------------------------------------------

# 0. Settup

```{r message=FALSE, warning=FALSE}
# Loading the needed libraries
library(dplyr)
library(here)
library(plm)
library(stringr)
library(readr)
library(ggplot2)
library(ggrepel)
library(lubridate)
library(sandwich)
library(modelsummary)
library(gt)
library(flextable)
library(broom)
library(texreg)
library(jtools)
library(estimatr)
library(parameters)

# Loading the database
ir_data<-read_csv(
          here("1_processed_data","database_for_IV_modeling.csv"), na = ""
          ) 

str(ir_data)

# Transformation of the dataframe to Paneldata
pdata <- pdata.frame(
          ir_data, index = c("country", "year")
          )
```

# 1. IV Assumptions: running F statistic test

```{r}
# First stage IV for rm_l3 instrument

IV1 <- iv_robust(gini ~ rm + gov_e + debt_s + ca_balance | 
                  rm_l3 + gov_e + debt_s + ca_balance, 
                  data = ir_data, diagnostics = TRUE)

msummary(IV1, stars = c('*' = .1, '**' = .05, '***' = .01))
```

```{r}
# Relevance rm_l3 test results

IV1$diagnostic_first_stage_fstatistic
scales::number(IV1$diagnostic_first_stage_fstatistic, accuracy = 0.01)
```

```{r}
# First stage IV for rm_l1 instrument

IV2 <- iv_robust(gini ~ rm + gov_e + debt_s + ca_balance | 
                  rm_l1 + gov_e + debt_s + ca_balance, 
                  data = ir_data, diagnostics = TRUE)

msummary(IV2, stars = c('*' = .1, '**' = .05, '***' = .01))
```

```{r}
# Relevance rm_l1 test results

IV2$diagnostic_first_stage_fstatistic
scales::number(IV2$diagnostic_first_stage_fstatistic, accuracy = 0.01)
```

```{r}
# First stage Over-identification test for both IVs

IV3 <- iv_robust(gini ~ rm + gov_e + debt_s + ca_balance | 
                  rm_l1 + rm_l3 + gov_e + debt_s + ca_balance, 
                  data = ir_data, diagnostics = TRUE)

msummary(IV3, stars = c('*' = .1, '**' = .05, '***' = .01))
```

```{r}
# Over-identification test results:
# Cannot reject that the two diff effects are the same. No evidence of diff coeff. 
# We can include both instrumets as instrumental variables. 

IV3$diagnostic_overid_test
scales::number(IV3$diagnostic_overid_test, accuracy = 0.01)
```

```{r}
# First is important to estimate the first stage of the model 
# Therefore, regress the treatment/endogenous variable on the controls and instruments 
# (Method extracted from https://www.youtube.com/watch?v=9IIGTk_AH9c)

FSmodels <- list(
                "FS1y" =  lm(rm~rm_l1+gov_e+ca_balance+debt_s, data=ir_data),
                "FS3y" = lm(rm~rm_l3+gov_e+ca_balance+debt_s, data=ir_data)
                )

modelsummary(FSmodels, stars=c('*' = .1, '**' = .05, '***' = .01)) #ADD , output = "FSIVmodels.docx"
```

```{r}
# Manually calculating the second stage for both instruments

FS1y <- lm(rm~rm_l1+gov_e+ca_balance+debt_s, data=ir_data)
FS3y <- lm(rm~rm_l3+gov_e+ca_balance+debt_s, data=ir_data)

set.seed(123)
rm_l1_hat <- predict(FS1y)
rm_l3_hat <- predict(FS3y)

FSIVmmodels <- list(
                    "FSIV 1y" =  lm(gini~rm_l1_hat+gov_e+ca_balance+debt_s, data=ir_data),
                    "FSIV 3y" = lm(gini~rm_l3_hat+gov_e+ca_balance+debt_s, data=ir_data)
                    )

modelsummary(FSIVmmodels, stars=c('*' = .1, '**' = .05, '***' = .01)) #ADD , output = "FSIVmodels.docx"
```

# 2. OLS Modeling

```{r warning=FALSE}
# Transformation of the dataframe to Paneldata

library(plm)
  
pdata <- pdata.frame(ir_data, index = c("country", "year"))

```

```{r}
# OLS Rm impact on inequality

OLSmodels <- list(
                  "OLS 1" = lm(gini ~ rm, data = pdata),
                  "OLS 2" = lm(gini ~ rm + gov_e + debt_s + ca_balance, data = pdata),
                  "OLS 1y" = lm(gini ~ rm_l1 + gov_e + debt_s + ca_balance, data = pdata),
                  "OLS 3y" = lm(gini ~ rm_l3 + gov_e + debt_s + ca_balance, data = pdata)
                  )

msummary(OLSmodels, stars = c('*' = .1, '**' = .05, '***' = .01))
```

```{r}
# Applying fixed effects at a year and country level
  
FEmodels <- list(
                  "FE Current Rm" = plm(gini ~ rm + gov_e + debt_s + ca_balance + factor(country), 
                                    data = pdata, model = "within", index = c("country")),
                  "FE Rm 1y" = plm(gini ~ rm_l1 + gov_e + debt_s + ca_balance + factor(country), 
                                   data = pdata, model = "within", index = c("country")),
                  "FE Rm 3y" = plm(gini ~ rm_l3 + gov_e + debt_s + ca_balance + factor(country), 
                                    data = pdata, model = "within", index = c("country"))
                )
  
msummary(FEmodels, stars = c('*' = .1, '**' = .05, '***' = .01))
```

# 3. IV Modeling: Testing two null hypoteses

```{r message=FALSE}
#Attaching packages that will be used for calculating 2SLS
library(tidyverse)
library(parameters)
library(modelsummary)
library(fixest)
```

## 3.1. 2SLS models

-   **Instrument: 1-year lagged Remittances**

```{r}
#2SLS MODELING - instrument is rm_l1 

itest_rm_l1 <- feols(gini ~ gov_e + debt_s + 
            factor(country) | rm ~ rm_l1, 
            cluster = ~incomegroup, data = ir_data)

msummary(list('First Stage' = itest_rm_l1$iv_first_stage[[1]],
                'Second Stage' = itest_rm_l1),
                coef_rm_l1 = c(
                                rm_l1 = '1 Year lagged Remmitances',
                                fit_rm = 'Remmitances'),         
                stars = c('*' = .1, '**' = .05, '***' = .01))
```

-   **Instrument: 3-year lagged Remittances**

```{r}

#2SLS MODELING - intrument is rm_l3 
itest_rm_l3 <- feols(gini ~ gov_e + debt_s + ca_balance +
            factor(country) | rm ~ rm_l3, 
            cluster = ~incomegroup, data = ir_data)

msummary(list('First Stage' = itest_rm_l3$iv_first_stage[[1]],
                'Second Stage' = itest_rm_l3),
                coef_rm_l3 = c(rm_l3 = '3 Year lagged Remmitances',
                fit_rm = 'Remmitances'),         
                stars = c('*' = .1, '**' = .05, '***' = .01),
         output = "S2LSVmodel_rm3.docx")
```

## 3.2. GMM models

```{r}
# Specific libraries needed for this estimations
library(modelsummary)
library(gmm)
library(sandwich)
```

-   **y= gini \| Instruments: 1 and 3-years lagged Remittances \| Testing Ho1**

```{r}
#GMM MODELING - y= gini, instrument are rm_l1 and rm_l3
GMMMmodel <- gmm(gini ~ gov_e + ca_balance + debt_s +
                  factor(country) + rm, 
                  ~ gov_e + ca_balance + debt_s +
                  factor(country) + rm_l3 + rm_l1, data = ir_data
                  )

#Applying clustering at income group level
msummary(GMMMmodel, vcov = ~incomegroup, stars = c('*' = .1, '**' = .05, '***' = .01))
```

# 4. Robustness Test and Extended Analysis

```{r}
# Load important library
library(lmtest)
```

## 4.1 Validity Tests

```{r}
# Running the Durbin-Wu-Hausman test GMM related dto Ho1

gmm_gini_validity <- lm(gini ~ gov_e + ca_balance + debt_s + factor(country) + rm, data = ir_data)


durbin_wu_hausman_test_gmm_gini <- dwtest(gmm_gini_validity, alternative = "greater")


print(durbin_wu_hausman_test_gmm_gini)
```

```{r}
# Running the Durbin-Wu-Hausman test for the GMM related to Ho2

gmm_h2_validity <- lm(palma ~ debt_s + rm + ca_balance + factor(country) + gov_e, data = ir_data)


durbin_wu_hausman_test_gmm_h2 <- dwtest(gmm_h2_validity, alternative = "greater")


print(durbin_wu_hausman_test_gmm_h2)

```

## 4.2 Different Dependent Variables

```{r}
# GMM MODELING - y= palma, instrument are rm_l1 and rm_l3
# Palma index= β_0+ β_1 (Rm) ̂+β_2 Control variables+ε                                                    (5.1)

GMMMmodel_palma <- gmm(palma ~ gov_e + ca_balance + debt_s +
                        factor(country) + rm, 
                        ~ gov_e + ca_balance + debt_s +
                        factor(country) + rm_l3 + rm_l1, data = ir_data)

#GMM MODELING - y= theil, instrument are rm_l1 and rm_l3
#Theil index= β_0+ β_1 (Rm) ̂+β_2 Control variables+ε                                                     (5.2)

GMMMmodel_theil <- gmm(theil ~ gov_e + ca_balance + debt_s +
                        factor(country) + rm, 
                        ~ gov_e + ca_balance + debt_s +
                        factor(country) + rm_l3 + rm_l1, data = ir_data)



# Summary table for the 3 GMM models (gini, palma and theil) 
msummary(list(GMMMmodel, GMMMmodel_palma, GMMMmodel_theil), 
         vcov = ~incomegroup, 
         stars = c('*' = .1, '**' = .05, '***' = .01))
```

```{r}
# Performing Variables transformations to test Ho2
#  Generating new lagged variables
library(Hmisc)

ire_data <- ir_data  %>%
            mutate(Lag(debt_s, -1)) %>%
            mutate(Lag(debt_s, -3))

# Consolidating new dataframe with new instrumental variables and renaming proceure 
ire_data <- ire_data  %>%
            rename(debt_s_l1 = `Lag(debt_s, -1)`) %>%
            rename(debt_s_l3 = `Lag(debt_s, -3)`) 

head(ire_data)
```

-   **y= palma \| Instruments: 1 and 3-years lagged debt \| Testing Ho1**

```{r}
# GMM MODELING - y= palma, instrument are debt_s_l1 and debt_s_l3

# Gov Exp= γ_0+ γ_1 〖Debt Service〗\_(1 or 3 years lagged)+γ_2 Control variables+ υ\
# Palma index= β_0+ β_1 (Gov Exp) ̂+β_2 Control variables+ ε

ire_data <- ire_data %>%
            select(palma, gini, gov_e, ca_balance, rm, debt_s, year, 
            country, incomegroup, debt_s_l1, debt_s_l3) %>%
             na.omit()


GMMMmodel_h2 <- gmm(palma ~ ca_balance + rm +
                    factor(country) + gov_e, 
                    ~ gov_e + ca_balance + rm +
                    factor(country) + debt_s_l1 + debt_s_l3, data = ire_data)

#Applying clustering at income group level
msummary(list(GMMMmodel_palma, GMMMmodel_h2), vcov = ~incomegroup, 
         stars = c('*' = .1, '**' = .05, '***' = .01))
```


